Legion: Expressing Locality and Independence with Logical Regions

### 2012-05-04 doc/sc2012/soop.tex
soop

Legion uses a {\em  software out-of-order processor}, or SOOP, to schedule tasks.  The SOOP 
is pipelined, distributed and extracts nested parallelism from subtasks.
There are two major challenges in implementing an efficient task scheduler:
\begin{itemize}
\item  For correctness, Legion must preserve data dependences between tasks.

@@

To solve the second problem Legion uses a {\em deferred execution model} that decouples the issuing
of operations from when operations are performed.  An issued operation waits for other operations on
which it is dependent to complete before executing, but a waiting operation does not cause the SOOP
to block.  With deferred execution, it is worthwhile for the SOOP to run (potentially far) ahead of actual execution,
allocating resources to and issuing tasks that must wait because they have data dependences on earlier tasks that have not 
finished executing.   Pipelining and distributing the SOOP further improves throughput and hides latency.

To solve the first problem, detecting and enforcing data dependences
efficiently at run time, the SOOP considers, for each task $t$, the
the regions $t$ uses and ensures that $t$ waits on any
earlier task whose region uses may conflict with $t$'s.  There is a
difficulty, however.  The programmer's regions are logical; a region
names a set of objects, but does not say where those objects
are located.  To ensure dependences are satisfied the SOOP must also {\em map} $t$: assign an
appropriate physical instance to each of $t$'s logical region.  Thus, dependence
analysis requires knowing both which tasks $t$ depends on and how those
other tasks are mapped.

Our SOOP design solves this problem by dividing the analysis of
dependences between two pipeline stages.  The first SOOP stage
computes task dependences at the granularity of logical regions
(Section~\ref{sec:dep}), which does not give enough information to map
$t$, but does identify other tasks that must map before $t$ can map.
The third SOOP stage maps $t$ by carrying out a more refined analysis
of the (already mapped) tasks on which $t$ depends
(Section~\ref{sec:map}).  Between these stages, the second stage
distributes tasks to other processors (Section~\ref{sec:dist}).  Once
a task has been assigned a processor and physical instances it is
issued for deferred execution (Section~\ref{sec:exec}).  The final
SOOP stage reclaims resources from completed tasks
(Section~\ref{sec:clean}).

@@

When a Legion task runs, it may spawn new subtasks to be executed.  
These subtasks are {\em children} of the {\em parent} task and {\em siblings} of each other.  
A task always executes on a single processor, but
its subtasks may execute elsewhere.

Each processor in the system runs an instance of the SOOP.
When a parent task spawns a child task, the child is {\em registered}
with the SOOP on the parent's processor;
registration records the subtask's logical regions,
privileges and coherence properties.  Children are
registered in the sequential order the parent
spawns them and enqueued for mapping dependence analysis.  In the
circuit simulation in Listing~\ref{lst:code_ex}, the spawn
statements on lines 32-34 register all three
kinds of subtasks (in program order) on the processor where {\tt
simulate\_circuit} executes.

Detecting mapping dependences between a newly registered task $t$ and
a previously registered task $t'$ requires comparing the two sets of
logical regions accessed.  For each logical region used by $t'$ that
may overlap with a logical region used by $t$, the privileges and
coherence modes are compared to determine whether a dependence exists.
If both regions need only read privileges there is never a dependence,
but if either task needs write or reduction privileges, the coherence
modes are compared using the table in Figure~\ref{fig:dependence}.
{\tt Dep} indicates dependence while {\tt None}
indicates independence.  {\tt Same} is a dependence unless the two tasks
use the same physical instance of the logical region.
%\footnote{For example, if two tasks wish to acess the region with atomic coherence, then if they are working on the same physical copy of the data atomicity can be guaranteed using, e.g., locking.  If they are working on different copies of the data then the only safe execution strategy for Legion is to serialize the two tasks.}
{\tt Cont} 
indicates a dependence unless a single writer has
atomic coherence.

@@

The table lists {\em simultaneous} and {\em relaxed} coherence modes
that we have not yet discussed.  Both modes
allow other tasks using the region to execute at the same time and differ
only in what data must be observed.  With simultaneous coherence, a task must 
see all updates to the logical region made by other tasks operating on the same region 
simultaneously (i.e., shared memory semantics).  With relaxed coherence, 
a task may or may not observe concurrent updates.

@@

A key property of Legion is that dependence analysis is only needed
between sibling tasks:
\begin{observation}
\label{obs:isolation}
\rm
Let $t_1$ and $t_2$ be two sibling tasks with no dependence.  Then no subtask of $t_1$ has a dependence with any subtask
of $t_2$.
\end{observation}
Recall from Section~\ref{sec:ex} that subtasks only access subregions of their parent task's region arguments (and with fewer
privileges).  Thus if the regions used by $t_1$ and $t_2$ do not overlap, the regions used by any subtasks of $t_1$ cannot
overlap with the regions used by any subtask of $t_2$.  If the regions used by $t_1$ and $t_2$ do overlap
but there is no dependence because the regions are used with simultaneous or relaxed coherence, then by definition there is no dependence between the subtasks.

Consider the first two subtasks spawned on line 32 of the circuit
simulation in Listing~\ref{lst:code_ex}, {\tt
calc\_new\_currents(pieces[0])} and {\tt
calc\_new\_currents(pieces[1])}.  The first of these tasks reads and
writes the private wires subregion {\tt pieces[0].rw\_pvt} in
exclusive mode, and reads in exclusive mode the node subregions {\tt
pieces[0].rn\_pvt}, {\tt .rn\_shr}, and {\tt
.rn\_ghost} (see line 39).  The second subtask must be
checked against the first for any dependences; the second subtask uses
{\tt pieces[1].rw\_pvt} (read/write exclusive) and {\tt
pieces[1].rn\_pvt}, {\tt .rn\_shr}, and {\tt .rn\_ghost} (read-only exclusive).
It may be helpful to refer to the region tree for nodes in Figure~\ref{sfig:part_fig:tree} (recall that the wires region tree consists of a single disjoint partition).  We give a few representative examples
of reasoning about pairwise dependences (not all pairs are covered):
\begin{enumerate}
\item {\tt pieces[0].rw\_pvt} and {\tt pieces[1].rw\_pvt} do not overlap as they are subregions of a disjoint partition.
%\item {\tt pieces[0].rn\_pvt} and {\tt pieces[1].rn\_pvt} do not overlap for the same reason.
%\item {\tt pieces[0].rn\_shr} and {\tt pieces[1].rn\_shr} do not overlap, also for the same reason.
\item {\tt pieces[0].rn\_ghost} may overlap with {\tt pieces[1].rn\_shr}, because they are in two different
  partitions of the same region.
%\item {\tt pieces[0].rn\_shr} and {\tt pieces[1].rn\_ghost} may overlap because they are in two different
  partitions of the same region.
\item {\tt pieces[0].rn\_ghost} may overlap with {\tt pieces[1].rn\_ghost} because they are in different subregions of
an aliased partition.
\item {\tt pieces[0].rw\_pvt} and {\tt pieces[1].rn\_pvt})
do not overlap because they are in different region trees and have no common region ancestor.
\end{enumerate}
Cases 2 and 3 indicate a possible dependence, however the regions are
accessed with only read privileges. All other cases for these two
subtasks are similar to one of the examples above; thus, the two
subtasks are independent.

A closer look at this example shows that whether there is overlap
between two regions $r_1$ and $r_2$ can be determined by examining
their least common ancestor $\lca{r_1}{r_2}$ in the region tree.  
There are four cases.  If $\lca{r_1}{r_2}$ either does not exist (the regions
are in different region trees) or is a disjoint partition, then $r_1$
and $r_2$ are disjoint.  If $\lca{r_1}{r_2}$ is an
aliased partition or a region, 
%(because $r_1$ and $r_2$ are in
%different partitions of the region, as in cases 4 and 5
%above), 
then $r_1$ and $r_2$ may not be disjoint.

@@

We can now describe the algorithm for the dependence analysis phase of
the SOOP.  The SOOP maintains a local region forest for each task $t$,
the roots of which are $t$'s region arguments and including all
partitions and subregions used by $t$.  Dependence analysis
places the child tasks of $t$ on the regions they use, maintaining the
following invariant. Let $r'$ be a region used by child task $t'$,
and let $R$ be the set of regions $r''$ such that $r''$ is used
by a task $t''$ registered after $t'$ and $t''$ depends on $t'$
because of overlap between $r''$ and $r'$.  Then $t'$ is listed on the
region that is the least common ancestor of the set $R \cup \{r'\}$.
Placing $t'$ on an ancestor of $r'$ has the effect of coarsening the
dependences involving $t'$, which
can result in false dependences.  However, the loss of
information can be shown to be small and only affects when $t$ can
map; any information lost is recovered during mapping.
Furthermore, the benefit is that it dramatically speeds up
dependence analysis overall.

@@

We illustrate dependence analysis using the four tasks discussed
above.  Many more tasks are executed by the circuit simulation,
but this subset illustrates the important
points. Figure~\ref{fig:depexamp} shows the node region tree in three
stages.  In Figure~\ref{fig:depexamp:a}, the two tasks
{\tt calc\_new\_currents(pieces[0])} ($c_0$) and {\tt
    calc\_new\_currents(pieces[1])} ($c_1$) have been placed on their region arguments,
  which determines these two tasks are independent.
Next, in Figure~\ref{fig:depexamp:b}, the task {\tt distribute\_charge(pieces[0],dt)} ($d_0$)
has dependences with both $c_0$ and $c_1$ caused by possible overlap between the pieces of the
shared node partition and the ghost node partition, so both $c_0$ and $c_1$ are moved to
the {\tt p\_nodes\_pvs[1]} region.  Finally, the task {\tt update\_voltages[1]} ($u_1$) causes
$d_0$ to also be moved to {\tt p\_nodes\_pvs[1]} for essentially the same reason; the final
state of the node region tree is shown in Figure~\ref{fig:depexamp:c}.

@@

\item If $t'$ has read/write privileges for $r'$, and $r, r'$ may
  overlap (i.e., $\lca{r}{r'}$ is a region or aliased
  partition), then $s'$ must be copied to an instance $s''$ of
  $\lca{r}{r'}$ and a fresh instance of $r'$ created from $s''$.
%  (if either $r = \lca{r}{r'}$ or $r' = \lca{r}{r'}$ one of the two
%  copies is omitted)
Instance $s'$ is removed from the region tree
  but not deallocated (see Section~\ref{sec:clean}).

\item If $t'$ has reduce privileges for $r'$ and $t$ has reduce
  privileges for $r$ and both use the same reduction operator, then
  nothing is done.  This is the only case in which a writer
  does not require a valid instance---because reductions can be
  reordered, combining the instances of $r$ and $r'$ (if they
  overlap) can be deferred to a later consumer of the data (see case
  4).

\item If $t'$ has reduce privilege for $r'$ and $t$ has read, write, 
  or reduce privileges with a different operator than $t'$
  for $r$, and $r$ and $r'$ may overlap, then $s'$ must be
  reduced (using $t'$'s reduction operator) to an instance $s''$ of
  $\lca{r}{r'}$ and then a fresh instance of $r'$ created from $s''$.
%  (If $r' = \lca{r}{r'}$ the reduction is unneeded, and if $r =
%  \lca{r}{r'}$ the copy is unneeded.) 
Instance $s'$ is removed from the region
  tree but not deallocated.
\end{enumerate}

>>>>

Legion uses a {\em  software out-of-order processor}, or SOOP, to schedule tasks.  A SOOP 
dynamically schedules a stream of tasks in a manner akin to an out-of-order processor scheduling a stream of instructions:
just as an instruction scheduler is constrained by register dependences,
the SOOP is constrained by region dependences.
The SOOP  is pipelined, distributed and extracts nested parallelism from subtasks.
There are two major challenges in implementing an efficient task scheduler:
\begin{itemize}
\item  For correctness, Legion must preserve data dependences between tasks, which is non-trivial
because the same data may be in multiple different regions or subregions.

@@

To solve the second problem Legion uses a {\em deferred execution model} that decouples the issuing
of operations from when operations are performed.  Using a low-level runtime event system (which we do
not describe further in this paper), an issued operation waits for other operations on
which it is dependent to complete before executing, but a waiting operation does not cause the SOOP
to block.  With deferred execution, it is worthwhile for the SOOP to run (potentially far) ahead of actual execution,
allocating resources to and issuing tasks that must wait because they have data dependences on earlier tasks that have not 
finished executing.   Pipelining and distributing the SOOP further improves throughput and hides latency.

To dynamically detect and enforce data dependences,
the SOOP considers, for each task $t$, the
the regions $t$ uses and ensures $t$ waits on any
earlier task whose region uses conflict with $t$'s.  There is a
difficulty, however.  The programmer's regions are logical; a region
names a set of objects, but does not say where those objects
are in memory.  To ensure dependences are satisfied the SOOP must also {\em map} $t$: assign an
appropriate physical instance to each of $t$'s logical regions.  Thus, dependence
analysis requires knowing both which tasks $t$ depends on and how those
other tasks are mapped.

Our SOOP design solves this problem by dividing the analysis of
dependences between two pipeline stages.  The first SOOP stage
computes task dependences at the granularity of logical regions
(Section~\ref{sec:dep}), which does not give enough information to map
$t$, but does identify all tasks that must map before $t$ can map.
The third SOOP stage maps $t$ by carrying out a more refined analysis
of the (already mapped) tasks on which $t$ depends
(Section~\ref{sec:map}).  Between these stages, the second stage
distributes tasks to other processors (Section~\ref{sec:dist}).  Once
a task has been assigned a processor and physical instances it is
issued for deferred execution (Section~\ref{sec:exec}).  The final
SOOP stage reclaims resources from completed tasks
(Section~\ref{sec:clean}).

@@

%When a Legion task runs, it may spawn new subtasks to be executed.  
%These subtasks are {\em children} of the {\em parent} task and {\em siblings} of each other.  
%A task always executes on a single processor, but
%its subtasks may execute elsewhere.

Each processor in the system runs an instance of the SOOP.
When a {\em parent} task spawns a {\em child} task, the child is {\em registered}
with the SOOP on the parent's processor;
registration records the subtask's logical regions,
privileges and coherence properties.  Children are
registered in the sequential order the parent
spawns them and enqueued for mapping dependence analysis.  In the
circuit simulation in Listing~\ref{lst:code_ex}, the spawn
statements on lines 32-34 register all three
kinds of subtasks (in program order) on the processor where {\tt
simulate\_circuit} executes.

Detecting mapping dependences between a newly registered task $t$ and
a previously registered task $t'$ requires comparing the two sets of
logical regions accessed.  For each logical region used by $t'$ that
may {\em alias} (may share data with) a logical region used by $t$, the privileges and
coherence modes are compared to determine whether a dependence exists.
If both regions need only read privileges there is never a dependence,
but if either task needs write or reduction privileges, the coherence
modes are compared using the table in Figure~\ref{fig:dependence}.
{\tt Dep} indicates dependence while {\tt None}
indicates independence.  {\tt Same} is a dependence unless the two tasks
use the same physical instance of the logical region.
%\footnote{For example, if two tasks wish to acess the region with atomic coherence, then if they are working on the same physical copy of the data atomicity can be guaranteed using, e.g., locking.  If they are working on different copies of the data then the only safe execution strategy for Legion is to serialize the two tasks.}
{\tt Cont} 
indicates a dependence unless a single writer has
atomic coherence.

@@

The table lists {\em simultaneous} and {\em relaxed} coherence modes
that we have not yet discussed.  Both modes
allow other tasks using the region to execute at the same time and differ
only in what updates must be observed.  With simultaneous coherence, a task must 
see all updates to the logical region made by other tasks operating on the same region 
simultaneously (i.e., shared memory semantics).  With relaxed coherence, 
a task may or may not observe concurrent updates.

@@

A key property of Legion is that dependence analysis is only needed
between sibling tasks:
\begin{observation}
\label{obs:isolation}
\rm
Let $t_1$ and $t_2$ be two sibling tasks with no dependence.  Then no subtask of $t_1$ has a dependence with any subtask
of $t_2$.
\end{observation}
Recall from Section~\ref{sec:ex} that subtasks only access subregions of their parent task's region arguments (and with fewer
privileges).  Thus if the regions used by $t_1$ and $t_2$ do not alias, the regions used by any subtasks of $t_1$ cannot
alias the regions used by any subtask of $t_2$.  If the regions used by $t_1$ and $t_2$ alias
but there is no dependence because the regions are used with simultaneous or relaxed coherence, then by definition there is no dependence between the subtasks.

Consider the first two subtasks spawned on line 32 of the circuit
simulation in Listing~\ref{lst:code_ex}, {\tt
calc\_new\_currents(pieces[0])} and {\tt
calc\_new\_currents(pieces[1])}.  The first of these tasks reads and
writes the private wires subregion {\tt pieces[0].rw\_pvt} in
exclusive mode, and reads in exclusive mode the node subregions {\tt
pieces[0].rn\_pvt}, {\tt .rn\_shr}, and {\tt
.rn\_ghost} (see line 39).  The second subtask must be
checked against the first for any dependences; the second subtask uses
{\tt pieces[1].rw\_pvt} (read/write exclusive) and {\tt
pieces[1].rn\_pvt}, {\tt .rn\_shr}, and {\tt .rn\_ghost} (read-only exclusive).
It may be helpful to refer to the region tree for nodes in Figure~\ref{sfig:part_fig:tree} (recall that the wires region tree consists of a single disjoint partition).  We give a few representative examples
of reasoning about pairwise dependences (not all pairs are covered):
\begin{enumerate}
\item {\tt pieces[0].rw\_pvt} and {\tt pieces[1].rw\_pvt} do not alias as they are subregions of a disjoint partition.
%\item {\tt pieces[0].rn\_pvt} and {\tt pieces[1].rn\_pvt} do not alias for the same reason.
%\item {\tt pieces[0].rn\_shr} and {\tt pieces[1].rn\_shr} do not alias, also for the same reason.
\item {\tt pieces[0].rn\_ghost} aliases {\tt pieces[1].rn\_shr}, because they are in two different
  partitions of the same region.
%\item {\tt pieces[0].rn\_shr} and {\tt pieces[1].rn\_ghost} alias because they are in two different
  partitions of the same region.
\item {\tt pieces[0].rn\_ghost} aliases {\tt pieces[1].rn\_ghost} because they are in different subregions of
an aliased partition.
\item {\tt pieces[0].rw\_pvt} and {\tt pieces[1].rn\_pvt})
do not alias because they are in different region trees and have no common region ancestor.
\end{enumerate}
Cases 2 and 3 indicate a possible dependence, however the regions are
accessed with only read privileges. All other cases for these two
subtasks are similar to one of the examples above; thus, the two
subtasks are independent.

A closer look at this example shows that whether there is aliasing
between two regions $r_1$ and $r_2$ can be determined by examining
their least common ancestor $\lca{r_1}{r_2}$ in the region tree.  
There are four cases.  If $\lca{r_1}{r_2}$ either does not exist (the regions
are in different region trees) or is a disjoint partition, then $r_1$
and $r_2$ are disjoint.  If $\lca{r_1}{r_2}$ is an
aliased partition or a region, 
%(because $r_1$ and $r_2$ are in
%different partitions of the region, as in cases 4 and 5
%above), 
then $r_1$ and $r_2$ may not be disjoint.

@@

We can now describe the algorithm for the dependence analysis phase of
the SOOP.  The SOOP maintains a local region forest for each task $t$,
the roots of which are $t$'s region arguments and including all
partitions and subregions used by $t$.  Dependence analysis
places the child tasks of $t$ on the regions they use, maintaining the
following invariant. Let $r'$ be a region used by child task $t'$,
and let $R$ be the set of regions $r''$ such that $r''$ is used
by a task $t''$ registered after $t'$ and $t''$ depends on $t'$
because of aliasing between $r''$ and $r'$.  Then $t'$ is listed on the
region that is the least common ancestor of the set $R \cup \{r'\}$.
Placing $t'$ on an ancestor of $r'$ has the effect of coarsening the
dependences involving $t'$, which
can result in false dependences.  However, the loss of
information can be shown to be small and only affects when $t$ can
map; any information lost is recovered during mapping.
Furthermore, the benefit is that it dramatically speeds up
dependence analysis overall.

@@

We illustrate dependence analysis using the four tasks discussed
above.  Many more tasks are executed by the circuit simulation,
but this subset illustrates the important
points. Figure~\ref{fig:depexamp} shows the node region tree in three
stages.  In Figure~\ref{fig:depexamp:a}, the two tasks
{\tt calc\_new\_currents(pieces[0])} ($c_0$) and {\tt
    calc\_new\_currents(pieces[1])} ($c_1$) have been placed on their region arguments,
  which determines these two tasks are independent.
Next, in Figure~\ref{fig:depexamp:b}, the task {\tt distribute\_charge(pieces[0],dt)} ($d_0$)
has dependences with both $c_0$ and $c_1$ caused by aliasing between the pieces of the
shared node partition and the ghost node partition, so both $c_0$ and $c_1$ are moved to
the {\tt p\_nodes\_pvs[1]} region.  Finally, the task {\tt update\_voltages[1]} ($u_1$) causes
$d_0$ to also be moved to {\tt p\_nodes\_pvs[1]} for essentially the same reason; the final
state of the node region tree is shown in Figure~\ref{fig:depexamp:c}.

@@

\item If $t'$ has read/write privileges for $r'$, and $r$ aliases $r'$,
   then $s'$ must be copied to an instance $s''$ of
  $\lca{r}{r'}$ and a fresh instance of $r'$ created from $s''$.
%  (if either $r = \lca{r}{r'}$ or $r' = \lca{r}{r'}$ one of the two
%  copies is omitted)
Instance $s'$ is removed from the region tree
  but not deallocated (see Section~\ref{sec:clean}).

\item If $t'$ has reduce privileges for $r'$ and $t$ has reduce
  privileges for $r$ and both use the same reduction operator, then
  nothing is done.  This is the only case in which a writer
  does not require a valid instance---because reductions can be
  reordered, combining the instances of $r$ and $r'$ (if they
  alias) can be deferred to a later consumer of the data (see case
  4).

\item If $t'$ has reduce privilege for $r'$ and $t$ has read, write, 
  or reduce privileges with a different operator than $t'$
  for $r$, and $r$ aliases $r'$, then $s'$ must be
  reduced (using $t'$'s reduction operator) to an instance $s''$ of
  $\lca{r}{r'}$ and then a fresh instance of $r'$ created from $s''$.
%  (If $r' = \lca{r}{r'}$ the reduction is unneeded, and if $r =
%  \lca{r}{r'}$ the copy is unneeded.) 
Instance $s'$ is removed from the region
  tree but not deallocated.
\end{enumerate}
